\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-12-08 response_1/response_1.tex

5.1. A 10 day generation interval is assumed based on a 20 day infectious period, however this is strongly dependent on assumptions about how infectiousness varies over the course of infection. Note that I prefer the term `generation interval' for this particular situation, see eg https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2365921/. In a simple SIR model, the mean generation interval will equal the mean infectious period, and the choice of 10 days seems a little strange. There is certainly considerable uncertainty about parameters here, so at least the authors could compute this over a range of values.

\textbf{Thank you for pointing this out. This was an error on our part. We've switched from a 10 day generation interval to a 20 day generation interval consistent with existing data and a simple SIR model.}

5.2. Other theoretical work suggests a slightly different relationship between Ne tau and I(t), specifically that it will depend more on transmission rate than generation time. However if making the approximation that $I(t)$ is constant (which the authors are) , the formulas will be the same up to a factor of 2$\times$. See Koelle, 2011, DOI: 10.1098/rsif.2011.0495 \& Volz and Frost 2013, DOI: 10.1098/rstb.2010.0060. Specifically, I would use $N_e\tau = I(t) \, / \, 2 \, \times $ [per capita transmission rate], but some of the formulas in Koelle's paper may be more realistic for MERS.
and equating $\beta = 1 \, / $ [generation time] gives 2$\times$ the authors' estimate if solving for $I$.
Note that if using a 20 day generation interval and the above formula, you would have $I = 127$, so this doesn't make any practical difference, but is more elegant in my opinion, since we are exact about the epidemic model that the estimate is based on. If some estimate of the dispersion parameter for $R_0$ could be obtained from the cluster size distribution, a better estimate for $I$ could be obtained using equations in Koelle 2011.

\textbf{This was helpful. We've reformulated this calculation to follow Frost and Volz (2010). We believe this reformulation streamlined the description. However, the switch from 10 day generation interval to 20 day interval resulted in a 2$\times$ difference to our final estimate. However, the final result of $N_{e}\tau=280.4$ years is still consistent with broad conclusions of a congruence (or at least no worse a congruence than measles in humans) between camel incidence rates and observed $N_e\tau$. Relevant text in Results has been reworked:}

### 2017-12-11 response_1/response_1.tex

Original comments are in plain text.  Our responses follow in \textbf{bold}.

\section*{Essential revisions}
1. The population genetic model (the particular form of structured coalescent) is highly idealised and this may influence the quantitative conclusions, although we suspect the conclusions are quite robust qualitatively. This model specifically estimates the rate of a lineage moving between demes going backwards in time; the numbers cited for the camel$\rightarrow$human rate is really the rate that a lineage in humans goes to a camel going down the tree. The relationship between these migration rates and the epidemiologically meaningful transmission rate is complex and depends among other things on the ratio of population sizes in both demes. Per-capita transmission rates could be estimated using an epidemiologically structured coalescent model (see e.g.\ papers by Volz \& Rasmussen), which would ideally be stochastic due to bursty dynamics in humans. But this would be a large undertaking and so we suggest that for now the distinction is clarified. Overall, a little more discussion of the complexity and pitfalls when relating idealised population genetic models (like the island model used here) to a noisy nonlinear epidemic like this one might be merited.

\textbf{Yes, we agree that the structured coalescent approach is idealised and does not reflect a meaningful rate of zoonotic transfer of lineages, which is the reason we restricted any mention of rates to supplementary figures and do not attach numbers whenever rates are mentioned, but still report the number of introductions observed in the sequence data. We have altered Figure S2 in the reviewed manuscript to reflect that the rates shown are backwards in time. We have added the following sentences to the discussion to highlight the fact that the coalescent model employed is not ideal:}

\begin{quotation}
Although we recover migration rates from our model (Figure S2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
%\end{quotation}

[...]

%\begin{quotation}
Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate time varying migration rates and population sizes, integrate alternative sources of information and fit to stochastic nonlinear models (Rasmussen \textit{et al.}, 2014), which would be more appropriate for MERS-CoV.
Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data (K\"{u}hnert \textit{et al.}., 2016), but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
\end{quotation}

@@

\begin{quotation}
When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme.
This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
\end{quotation}

4. Easily addressed, but important. The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger. Antia et al.\ Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with $R_0>1$ increases dramatically as a function of the wild-type $R_0$. So $R_0\sim0.8$ is much worse than $R_0\sim0.3$. More sophisticated models have been done since, especially by Llyod-Smith's group, but the basic result is sound. In the light of this theoretical work, your findings are not at all reassuring.

\textbf{We agree that this is an important point, but also feel that it is difficult to formulate warnings about pandemic potential without overstating the case. We also believe that adaptive landscapes play a considerable role in emerging pandemics. We added an additional sentence to the discussion and refer to the Antia et al.\ study:}

\begin{quotation}
Previous modeling studies (Antia \textit{et al}., 2003; Park \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
\end{quotation}

@@

7. Page 9. I think the section on recombination is verbose, and potentially misleading. You didn't perform an ARG, you performed a perfectly reasonable sensitivity analysis to different partitions of the genome. I would be more upfront about this, and try to explain it more simply and clearly.

\textbf{We have tried our best to clean up the section on recombination.}

8. Discussion. I think the results of this analysis on $R_0$ and cluster size distribution could be compared with Cauchemez et al.\ PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters. An important result of that paper was that the cluster size distributions were very skew - I think it would be good to look at skewness here.

\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al}. study did not share any data beyond the raw input file they used. Since neither the actual numbers in Fig 1C of Cauchemez \textit{et al}., nor an implementation of the MCMC algorithm exist in the public domain we have included an additional figure (supplementary Figure S8) that visualises the distribution of cases recovered by simulation. These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al}.}

### 2017-12-12 mers-structure.tex
Reduce abstract to ~150 words.

Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
While case-based and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that human outbreaks in the Arabian peninsula are driven by seasonally varying zoonotic transfer of viruses from camels.
Without heretofore unseen evolution of host tropism, MERS-CoV is unlikely to become endemic in humans.

>>>>

Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus from camels causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, and while case-based and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential.
Here we use existing MERS-CoV sequence data to explore its phylodynamics in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that human outbreaks in the Arabian peninsula are driven by seasonally varying zoonotic transfer of viruses from camels.
Without heretofore unseen evolution of host tropism, MERS-CoV is unlikely to become endemic in humans.
